---
title: "Predicting Weight Lifting Performance"
author: "Cynthia S Chan"
date: "November 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
The following analysis intends to build a prediction model on the manner in which participants did weight lifting exercises.  The data comes from accelerometers on the belt, forearm, arm, and dumbbell of six participants while they perform barbell lifts correctly and incorrectly in 5 different ways. 


## Data Processing
The data for the project come from this source http://groupware.les.inf.puc-rio.br/har.  There is a set of training and a set of testing data.   

```{r data, echo=TRUE}
##  download file by url from internet to working directory

##  training data
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile="pml-training.csv")

##  testing data
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile="pml-testing.csv")

##  read in data
training <- read.csv("./pml-training.csv", header=TRUE, na.strings=c("NA", "#DIV/0!"))
testing <- read.csv("./pml-testing.csv", header=TRUE, na.strings=c("NA", "#DIV/0!"))
dim(training); dim(testing)

```


The training data is split so that one part is used to pick the features and train the prediction function, and the other part s used to cross-validate the prediction model. First the data is tidied up to remove variable columns that have a lot of missing values or will not be used as predictors.

```{r split, echo=TRUE}
##  load packages
library(caret)
library(MASS)
library(rpart)
library(rattle)
library(e1071)

##  tidy data
training <- training[, colSums(is.na(training))==0]
training <- training[, -c(1:7)]
dim(training)

##  split for training and cross-validation
set.seed(123)
inTrain <- createDataPartition(training$classe, p=3/4, list=FALSE)
traindata <- training[inTrain,]
validata <- training[-inTrain,]
```

##  Data pre-processing
There are now 53 variables in the training data.  The near zero function is used to identify any variables that have very little variability and are not likely to be useful predictors.

```{r preproc, echo=TRUE}
nsv <- nearZeroVar(traindata, saveMetrics = TRUE)
table(nsv$zeroVar, nsv$nzv)
```

##  Prediction Algorithms
There aren't any variables with little to no variability, so all will be used to predict the classe outcome.
First, linear discriminant analysis (LDA) is tested to look for possible linear combinations of the continuous accelerometer measurements that may distinguish the classes of exercise.

```{r lda, echo=TRUE}
modLDA <- train(classe ~., data=traindata, method="lda")
predLDA <- confusionMatrix(predict(modLDA, validata), validata$classe)
predLDA
```

The LDA method resulted in moderate classification (70% accuracy), thus it may be that the relationship between the predictive factors and variables are more complex or non-linear.  Next, a classification tree is attempted.  

```{r tree, echo=TRUE}
modTree <- train(classe ~., data=traindata, method="rpart")
predTree <- confusionMatrix(predict(modTree, validata), validata$classe)
predTree

fancyRpartPlot(modTree$finalModel)
```

The classification tree resulted in poor accuracy (<50%), and as indidcated by the sensitivity results and demonstrated in the tree plot, it is unable to distinguish classe D.

Instead a support vector machine (SVM) algorithm is tried.


```{r svm, echo=TRUE}
modSVM <- svm(classe ~., data=traindata)
predSVM <- confusionMatrix(predict(modSVM, validata), validata$classe)
predSVM
```

The SVM model resulted in correctly classifying 94% of the validation data set, and has high sensitivity and specificify across all 5 classes.  With an error rate of approximately 6% on the validation data that was not used for training, it is expected that this model will generalize well to new data and the out of sample rate will be similar.  


This model will be tested on the test data set.  

```{r test, echo=TRUE}
prednew <- predict(modSVM, testing)
prednew
```



### References
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.
